Unsupervised embedding of trajectories captures the latent structure of mobility

### 2021-03-22 paper/rebuttal/rebuttal.tex

\newgeometry{margin=1in, top=1cm}
\begin{document}
\maketitle

\bigskip
\thispagestyle{empty}
\vspace{-0.4cm}
\noindent Dear Dr. Sutherland and Referees,\\
\vspace{-0.25cm}

We thank all the referees for their thoughtful and helpful comments and critiques.
We acknowledge that our initial submission had a substantial gap regarding the choice of the \textit{word2vec} model, its relationship to the gravity model of mobility, and its evaluation.
While reflecting on these comments and our manuscript, we have made a theoretical breakthrough that addresses the fundamental issues, and dramatically expands the scope of our contribution;
namely, we have discovered \emph{a mathematical equivalence between the \textit{word2vec} (skip-gram with negative sampling) model and the gravity model.}

Furthermore, our subsequent experiments show that our approach robustly outperforms exceeding many other baselines, even direct optimization of the gravity model.
Therefore, we are confident that these changes have brought a qualitative leap in the significance of our findings.
Thus, we would like to request a reconsideration of our manuscript.
Again, we are incredibly appreciative of the referees' comments, which provided the insight and motivation needed to make these breakthroughs.

@@

\vspace{-0.3cm}
\paragraph{Direct optimization of gravity model:} Finally, we create an embedding by optimizing for the gravity model directly.
Specifically, we construct a gravity matrix of all organizations, where each cell of the matrix is calculated as $T_{ij}/m_i m_j$, where $T_{ij}$ is the number of co-occurrences of organizations $i$ and $j$ in the same affiliation trajectory, and $m_i$ and $m_j$ are the organizations' populations, defined here as the mean annual number of unique mobile and non-mobile authors affiliated. We then embed the gravity matrix using two approaches: the truncated SVD, and multidimensional scaling (MDS), which embeds each location in an N-dimensional space such that pairwise distances are preserved as well as possible.
Typically, MDS uses Euclidean distance to measure the distance between vectors.

\subsection{New experimental parameters}
In addition to the new baselines, we have also incorporated additional experimental parameters into our analysis.
The first is $\gamma$, a hyper-parameter of the \textit{word2vec} model, which defines the exponent which shapes the negative sampling distribution.
The default of \texttt{gensim} implementation of \textit{word2vec} is set to $\gamma = 0.75$, and this was the value used in our original manuscript.
However, to connect \textit{word2vec} to the gravity model of mobility, we assume $\gamma = 1.0$ (see Section 1).
We now report all main results using $\gamma = 1.0$ and also compare its performance to embeddings learned using $\gamma = 0.75$, as we do with the number of embedded dimensions and window size.
We plot the performance of each combination of hyperparameters by their ability to explain real-world flux of scientific mobility (Fig.~\ref{fig:supp:hyperparameter}), and observe no noticeable difference in performance between values of $\gamma = 0.75$ and $\gamma = 1.0$.
Figures in the updated manuscript have been re-drawn with $\gamma = 1.0$, and the findings remain qualitatively similar.
The most notable differences are observed in figures 3 (the UMAP projection of all organizations) and 4 (the clustering of country-level vectors);
still, the findings from these plots are largely similar, and quantitative measurements associated with these plots robust. 
As a side benefit, including the hyperparameter of $\gamma$ also effectively doubles the number of embeddings trained, and that results are so similar across different embedding provides further evidence of the robustness of the approach applied to our data.

### 2021-03-23 paper/current/main.tex
small edits

In revealing the correspondence between neural embeddings and the gravity model, the study of human mobility can move beyond geographic and network-based models of mobility, and instead leverage the high-order structure from individuals' mobility trajectories.
While we focus on three domains of mobility, this approach could be applied to many different domains, such as general human migration, transit-network mobility, and more.
Once learned, functional distances between locations, such as countries, cities, or organizations, can be published to support future research.
Moreover, this approach can be used to learn a functional distance even between entities for which no geographic analog exists, such as between occupational categories based on individuals' career trajectories.
In addition to providing a functional distance that supports modeling and predicting mobility patterns, the structure of the neural embedding space is amenable to a range of unique applications for studying mobility.
As we have shown, the embedding space allows the visualization of the complex structure of scientific mobility at high resolution across multiple scales, providing a large and detailed map of the landscape of global scientific mobility.
Embedding also allows us to quantitatively explore abstract notions such as academic prestige, and can potentially be generalized to other abstract axes.
Investigation of the structure of the embedding space, such as the vector norm, reveals universal patterns based on the organization's size and their vector norm that could be leveraged in future studies of mobility.

This approach, and our study, also have several limitations.
First, the skip-gram \textit{word2vec} model does not leverage directionality, meaning that embedding will be less effective at capturing mobility for which directionality is critical.
Future studies may consider bi-directional embeddings, such as BERT~\autocite{devlin2018bert}, to incorporate directionality, as well as their correspondence to asymmetric mobility models, such as the radiation model~\cite{simini2012universal}.
Second, the neural embedding approach is most useful in cases of mobility between discrete geographic units such as between countries, cities, and businesses; this approach is less useful in the case of mobility between locations represented using geographic coordinates, such as in the modeling of animal movements.
Third, neural embeddings are an inherently stochastic procedure, and so results may change across different iterations.
However, in this study we observe all results to be robust to stochasticity, likely emerging from the limited ``vocabulary'' of scientific mobility, airports, and accommodations (several thousand) and the relatively massive datasets used to learn representations (several million trajectories).
Applications of \textit{word2vec} to problem domains where the ratio of the vocabulary to data is smaller, however, should be implemented with caution to ensure that findings are not the result of random fluctuations.
Finally, the case of scientific mobility presents domain-specific limitations.
Reliance on bibliometric metadata means that we capture only long-term mobility such as migration, rather than the array of more frequent short-term mobility such as conference travel and temporary visits.
The kinds of mobility we do capture---migration and co-affiliation---although conceptually different, are treated identically by our model.
Also, our data might further suffer from bias based on publication rates: researchers at prestigious organizations tend to have more publications, leading to these organizations appearing more frequently in affiliation trajectories.

@@

Negative sampling trains {\it word2vec} as follows.
For each target word $i$, we sample a context word $j$ from the given data and label it as positive, denoted by $Y_{j}=1$.
Then, we sample $k$ words $\ell$ from a noise distribution $p_0(\ell)$ and label them as negative, denoted by $Y_{\ell}=0$.
In \textit{word2vec}, the noise distribution is given by $p_0(\ell) \propto P^\gamma (\ell)$, where $P(j)$ is the fraction of $j$ in the data, and $\gamma$ is a hyper-parameter.
Then, for the sampled words, we fit a logistic regression model
\begin{align}
	\label{eq:logistic-regress}
	P^{\text{NS}}(Y_{j} = 1; \vect{v}_{i}, \vect{u}_j) = \frac{1}{1 + \exp(-\vect{u}_j \cdot \vect{v}_{i})},
\end{align}
by maximizing the log-likelihood:
\begin{align}
	\label{eq:log-likelihood-logistic-regress}
	{\cal J}^{\text{NS}} = \sum_{i \in {\cal A}} \sum_{j \in {\cal D}} \left[ Y_{j} \log P^{\text{NS}}(Y_{j} = 1; \vect{v}_{i}, \vect{u}_j) + (1-Y_{j}) \log P^{\text{NS}}(Y_{j} = 0; \vect{v}_{i}, \vect{u}_j)\right],
\end{align}
where ${\cal D}$ is the set of all sampled context words.

@@

We compare the performance of the embedding distance to many baselines.
These include distances derived from simpler embedding approaches, such as Singular Value Decomposition (SVD) and a Laplacian Eigenmap embedding performed on the underlying location co-occurrence matrix.
We also use network-based distances, calculating vectors using a Personalized Page Rank approach and measuring the distance between them using cosine distance and Jensen-Shannon Divergence (see \SI).
Finally, we compare the embedding distance against embeddings calculated through direct matrix factorization, following the approach that \textit{word2vec} implicitly approximates~\autocite{levy2014neural}.

@@

We model co-occurrences $T_{ij}$ for locations $i$ and $j$ (referred to as flux), using the gravity law of mobility~\autocite{zipf1946gravity}.
The gravity law of mobility, which was inspired by Newton's law of gravity, postulates that attraction between two locations is a function of their population and the distance between them.
This formulation and variants have proven useful for modeling and predicting many kinds of mobility~\autocite{jung2008highwaygravity, curiel2018citygravity, truscott2012epidemicgravity, hong2016busgravity}.
In the gravity law of mobility, the \textit{expected flux}, $\hat{T}_{ij}$ between two locations $i$ and $j$ is defined as,

>>>>

In revealing the correspondence between neural embeddings and the gravity model, the study of human mobility can move beyond geographic and network-based models of mobility, and instead leverage the high-order structure from individuals' mobility trajectories using these robust and efficient methods. 
While we focus on three domains of mobility, this approach could be applied to many different domains, such as general human migration, transit-network mobility, and more.
Once learned, functional distances between locations, such as countries, cities, or organizations, can be published to support future research.
Moreover, this approach can be used to learn a functional distance even between entities for which no geographic analog exists, such as between occupational categories based on individuals' career trajectories. 
In addition to providing a functional distance that supports modeling and predicting mobility patterns, the structure of the neural embedding space is amenable to a range of unique applications for studying mobility.
As we have shown, the embedding space allows the visualization of the complex structure of scientific mobility at high resolution across multiple scales, providing a large and detailed map of the landscape of global scientific mobility.
Embedding also allows us to quantitatively explore abstract notions such as academic prestige, and can potentially be generalized to other abstract axes.
Investigation of the structure of the embedding space, such as the vector norm, reveals universal patterns based on the organization's size and their vector norm that could be leveraged in future studies of mobility.

This approach, and our study, also have several limitations.
First, the skip-gram \textit{word2vec} model does not leverage directionality, meaning that embedding will be less effective at capturing mobility for which directionality is critical.
Future studies may consider bi-directional embeddings, such as BERT~\autocite{devlin2018bert}, to incorporate directionality, as well as their correspondance to asymetric mobility models, such as the radiation model~\cite{simini2012universal}.
Second, the neural embedding approach is most useful in cases of mobility between discrete geographic units such as between countries, cities, and businesses; this approach is less useful in the case of mobility between locations represented using geographic coordinates, such as in the modeling of animal movements.
Third, neural embeddings are an inherently stochastic procedure, and so results may change across different iterations.
However, in this study we observe all results to be robust to stochasticity, likely emerging from the limited ``vocabulary'' of scientific mobility, airports, and accommodations (several thousand) and the relatively massive datasets used to learn representations (several million trajectories).
Applications of \textit{word2vec} to problem domains where the ratio of the vocabulary to data is smaller, however, should be implemented with caution to ensure that findings are not the result of random fluctuations.
Finally, the case of scientific mobility presents domain-specific limitations.
Reliance on bibliometric metadata means that we capture only long-term mobility such as migration, rather than the array of more frequent short-term mobility such as conference travel and temporary visits.
The kinds of mobility we do capture---migration and co-affiliation---although conceptually different, are treated identically by our model.
Also, our data might further suffer from bias based on publication rates: researchers at prestigious organizations tend to have more publications, leading to these organizations appearing more frequently in affiliation trajectories.

@@

Negative sampling trains {\it word2vec} using a binary classification task as follows.
For each target word $i$, we sample a context word $j$ from the given data and label it as positive, denoted by $Y_{j}=1$.
Then, we sample $k$ words $\ell$ from a noise distribution $p_0(\ell)$ and label them as negative, denoted by $Y_{\ell}=0$.
In \textit{word2vec}, the noise distribution is given by $p_0(\ell) \propto P^\gamma (\ell)$, where $P(j)$ is the fraction of $j$ in the data, and $\gamma$ is a hyper-parameter.
Then, for the sampled words, we fit a logistic regression model
\begin{align}
	\label{eq:logistic-regress}
	P^{\text{NS}}(Y_{j} = 1; \vect{v}_{i}, \vect{u}_j) = \frac{1}{1 + \exp(-\vect{u}_j \cdot \vect{v}_{i})},
\end{align}
by maximizing the log-likelihood:
\begin{align}
	\label{eq:log-likelihood-logistic-regress}
	{\cal J}^{\text{NS}} = \sum_{i \in {\cal A}} \sum_{j \in {\cal D}} \left[ Y_{j} \log P^{\text{NS}}(Y_{j} = 1; \vect{v}_{i}, \vect{u}_j) + (1-Y_{j}) \log P^{\text{NS}}(Y_{j} = 0; \vect{v}_{i}, \vect{u}_j)\right],
\end{align}
where ${\cal D}$ is the set of all sampled context words.

@@

We compare the performance of the embedding distance to many baselines.
These include distances derived from simpler embedding approaches, such as Singular Value Decomposition (SVD) and a Laplacian Eigenmap embedding performed on the underlying location co-occurrence matrix.
We also use network-based distances, calculating vectors using a Personalized Page Rank approach and measuring the distance between them using cosine distance and Jensen-Shannon Divergence (see \SI).
Finally, we compare the embedding distance against embeddings calculated through direct matrix factorization, following the approach that \textit{word2vec} implicitley approximates~\autocite{levy2014neural}.

@@

We model co-occurences $T_{ij}$ for locations $i$ and $j$ (referred to as flux), using the gravity law of mobility~\autocite{zipf1946gravity}.
The gravity law of mobility, which was inspired by Newton's law of gravity, postulates that attraction between two locations is a function of their population and the distance between them.
This formulation and variants have proven useful for modeling and predicting many kinds of mobility~\autocite{jung2008highwaygravity, curiel2018citygravity, truscott2012epidemicgravity, hong2016busgravity}.
In the gravity law of mobility, the \textit{expected flux}, $\hat{T}_{ij}$ between two locations $i$ and $j$ is defined as,
